---
title: "IdeoViz: a package for plotting simple data along ideograms"
author: "Shraddha Pai, Jingliang Ren" 
package: IdeoViz
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc_float: true
vignette: >
    %\VignetteIndexEntry{IdeoViz: a package for plotting simple data along ideograms} 
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

# IdeoViz: Plot data along chromosome ideograms
Plotting discrete or continuous dataseries in the context of chromosomal location has several useful applications in genomic analysis. Examples of possible metrics include RNA expression levels, densities of epigenetic marks or genomic variation, while applications could range from the analysis of a single variable in a single context, to multiple measurements in several biological contexts (e.g. age/sex/tissue/disease context). Visualization of metrics superimposed on the chromosomal ideogram could provide varied insights into the metric of interest:

* It could identify distinctive spatial distribution that could further hypotheses about the functional role of the metric (e.g. telocentric or pericentromeric enrichment)
* It could highlight distribution differences between different groups of samples, suggesting different regulatory mechanisms; in extreme cases, visualization may identify large genomic foci of differences
* It could confirm that a quantitative difference measured between groups of interest is consistent throughout the genome (i.e. that there are no foci, and that the change is global).

## Setup

```{r,eval=TRUE}
require(IdeoViz)
require(RColorBrewer) ### nice colours
data(binned_multiSeries)
```

## Example 1: Plotting several trendlines along one ideogram

The ideogram table containing cytogenetic band information is used to render chromosomes. This table corresponds directly to the `cytoBandIdeo` table from the UCSC genome browser. There are two ways to supply an ideogram table to `plotOnIdeo()`:

* First, it can be automatically downloaded from UCSC for your genome of choice, using the `getIdeo()` function. 
* Alternately, a pre-downloaded `cytoBandIdeo` table can be provided to downstream functions such as `plotOnIdeo()`. In this case, the table must be provided as a data.frame object with a header row and the column order matching that of the `cytoBandIdeo()` table at UCSC.

```{r,eval=TRUE}
ideo <- getIdeo("hg18")
head(ideo)
plotOnIdeo(chrom=seqlevels(binned_multiSeries), # which chrom to plot?
          ideoTable=ideo, # ideogram name
          values_GR=binned_multiSeries, # data goes here
          value_cols=colnames(mcols(binned_multiSeries)), # col to plot
          col=brewer.pal(n=5, 'Spectral'), # colours
          val_range=c(0,10), # set y-axis range
          ylab="array intensities",
          plot_title="Trendline example")  
```

## Example 2: Plotting a single series in bar format
For this example, we specify a local file to obtain the chromosome ideograms, rather than having IdeoViz download it from UCSC.

```{r,eval=TRUE}
data(binned_singleSeries)
data(hg18_ideo) # cytoBandIdeo table downloaded previously and stored as a data.frame.
plotOnIdeo(chrom=seqlevels(binned_singleSeries), 
          ideo=hg18_ideo,
          values_GR=binned_singleSeries, 
          value_cols=colnames(mcols(binned_singleSeries)),
          plotType='rect', # plot as bars
          col='blue', vertical=TRUE,
          val_range=c(-1,1), ylab="dummy score",
          plot_title="Discretized example")
```

## Example 3: Plotting a single series in bar format along entire genome
```{r,eval=TRUE}
data(binned_fullGenome)
plotOnIdeo(chrom=seqlevels(binned_fullGenome), 
          ideo=ideo, 
          values_GR=binned_fullGenome, 
          value_cols=colnames(mcols(binned_fullGenome)),
          plotType='rect',
          col='orange', 
          addScale=FALSE, # hide scale to remove visual clutter
          plot_title="Whole genome view",
          val_range=c(-1,1),cex.axis=0.5)
          #chromName_cex=0.6)
```

## Example 4: Binning data using IdeoViz functions
In this example, we do everything in IdeoViz: download the ideogram from UCSC, bin the data, and finally, plot along chromosomes.

For the example, we use histone H3K9me3 peak intensities mapped in the human lymphoblastoid cell line GM12878 (GEO accession GSM733664, only 3 chromosomes shown for simplicity). Here, average peak signal is plotted in 500Kb bins along the chromosome. The ideogram plots show high signal in pericentromeric and telomeric regions, consistent with the association of this histone mark with heterochromatin.

ENCODE Project Consortium, Bernstein BE, Birney E, Dunham I, Green ED, Gunter C, Snyder M. An integrated encyclopedia of DNA elements in the human genome. *Nature.*(2012): **489** (7414):57-74. 

```{r,eval=TRUE}
ideo_hg19 <- getIdeo("hg19")
chroms <- c("chr1","chr2","chrX")
data(GSM733664_broadPeaks)
head(GSM733664_broadPeaks)

chrom_bins <- getBins(chroms, ideo_hg19,stepSize=5*100*1000)
avg_peak <- avgByBin(data.frame(value=GSM733664_broadPeaks[,7]), 
  GSM733664_broadPeaks[,1:3], chrom_bins)

plotOnIdeo(chrom=seqlevels(chrom_bins), 
          ideoTable=ideo_hg19, 
          values_GR=avg_peak, value_cols='value', 
          val_range=c(0,50), 
          plotType='rect', 
          col='blue', vertical=TRUE
) 
```

## Example 5: Plotting a set of coordinates as tracks
Here we plot multiple GRanges(), each as its own track.
```{r,eval=TRUE}
ideo_hg19 <- getIdeo("hg19")
x <- GSM733664_broadPeaks
gr <- GRanges(x[,1],IRanges(x[,2],x[,3]))

pal <- brewer.pal(n=5,name="Dark2")
chroms <- c("chr1","chr2","chrX")
gr <- gr[which(seqnames(gr)%in% chroms)]
chrom_bins <- getBins(chroms, ideo_hg19,
					 stepSize=5*100*1000)
grList <- list(gr,gr,gr,gr,gr)
plotOnIdeo(chrom=seqlevels(chrom_bins), 
          ideoTable=ideo_hg19, 
          values_GR=grList, value_cols="value",  
          plotType="seg_tracks",
          col=pal, vertical=FALSE)
```

Segments can also be colour-coded by group type. For this the GRanges object needs to have a metadata column named "group", which has the pre-defined categories
```{r,eval=TRUE}
# assign group categories
for (k in 1:5) {
   gp <- rep("type1",length(grList[[k]])); 
   gp[(k*1000):((k*1000)+4000)] <- "type2"
   gp[1:1000] <- "type3" 
   grList[[k]]$group <- gp
   print(table(grList[[k]]$group))
}

# notice we don't name type3 - this is to show behaviour if a name is not specified
namedCols <- c("orange","purple"); names(namedCols) <- c("type1","type2")

plotOnIdeo(chrom=seqlevels(chrom_bins), 
    ideoTable=ideo_hg19,
    values=grList,
    plotType="seg_tracks",
    col=namedCols,
    vertical=FALSE)
```

### sessionInfo

```{r,eval=TRUE}
sessionInfo()
```